Semi-Permeable Vesicles: Mathematical Formulation, Numerics, and Physics

### 2020-06-04 vesicle_code/docs/permeableVes/main.tex
working on permeable paper

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section*{Examples}
We consider a single semi-permeable vesicle in four different flow
regimes. All vesicles have a length of $L=4\pi$, but the area is not
conserved because of semi-permeability. However, the area is bounded by
$4\pi$ which is the area of a circle with length $L=4\pi$. For each
simulation, we apply an SDC second-order adaptive time stepping
method~\cite{qua-bir2016}.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection*{Quiescent Flow} 
We consider a semi-permeable vesicle with initial reduced area $\nu =
0.65$ and several different permeability rates $\beta$ in a quiescent
flow. If the vesicle is impermeable, meaning that $\beta = 0$, it will
relax to a steady-state shape with negative and constant
tension~\cite{kra-win-sei-lip1996}. However, from
equation~\eqref{eqn:SSshape}, a semi-permeable vesicle relaxes to a
circle with positive constant tension. Since the area of a
semi-permeable vesicle is governed by~\eqref{eqn:areaROC}, we expect
both the time for the vesicle to deviate from its initial area, and the
time to reach its steady state will scale with $\beta^{-1}$. In
Figure~\ref{fig:ellipseArea}, we plot the area of a vesicle with
different permeability rates. The area is computed directly from the
vesicle shape (solid lines) and by integrating
equation~\eqref{eqn:areaROC} (marks) with a first-order method. We also
include black dashed lines indicating the times, $t_{\mathrm{steady}}$,
that the vesicle is near its steady state circular shape. In
Table~\ref{tbl:ellipseRelaxTimes}, we report the time,
$t_\mathrm{steady}$, required for the vesicle's area to be within 1.3\%
of its maximum area, and the time, $t_\mathrm{inflate}$, required for
the vesicle's area to exceed 1.3\% of its initial area. We see that both
times scale with $\beta^{-1}$.

\begin{table}[htp]
  \centering
  \caption{\label{tbl:ellipseRelaxTimes} The time for a semi-permeable
  vesicle to its steady state ($t_\mathrm{steady}$) and to begin
  inflating ($t_\mathrm{inflate}$).}
  \begin{tabular}{C{0.8cm}C{2cm}C{0.8cm}C{2cm}C{0.8cm}}
    $\beta$ & $t_\mathrm{steady}$ & ratio & $t_\mathrm{inflate}$ & ratio \\
    \midrule
    $10^{0}$  & $6.45 \times 10^{0}$ & ---  
              & $2.17 \times 10^{-2}$ & --- \\ 
    $10^{-1}$ & $2.00 \times 10^{1}$ & 3.10 & 
                $1.25 \times 10^{-1}$ & 5.76 \\
    $10^{-2}$ & $9.61 \times 10^{1}$ & 4.81 & 
                $1.59 \times 10^{0}$ & 12.72 \\
    $10^{-3}$ & $8.36 \times 10^{2}$ & 8.71 & 
                $2.19 \times 10^{1}$ & 13.77 \\
    $10^{-4}$ & $8.15 \times 10^{3}$ & 9.74 & 
                $2.14 \times 10^{2}$ & 9.77 \\
    $10^{-5}$ & $8.03 \times 10^{4}$ & 9.85 & 
                $2.20 \times 10^{3}$ & 10.14 \\
    \bottomrule
  \end{tabular}
\end{table}

@@

We further analyze the transient dynamics for small $\beta$ by
considering the vesicle velocity~\eqref{eqn:vesVelocity} due to
semi-permeability, $\beta(\ff \cdot \nn)\nn$, and force balance,
$\SS[\ff]$. Initially, since $\beta$ is small, the semi-permeability velocity is much
smaller than the balance velocity. Therefore,
the velocity resembles that of an impermeable vesicle. However, as the
vesicle approaches the steady state shape of an impermeable vesicle, the
force balance velocity decreases below the semi-permeability velocity,
at which point the semi-permeable vesicle begins to inflate.  Finally,
the vesicle semi-permeable inflates until it is circular with zero
membrane force, and therefore zero velocity.  Letting $\beta = 10^{-3}$,
in Figure~\ref{fig:vesVelocity} we plot the force balance velocity
(blue) and the semi-permeability velocity (red) at three times. We also
plot the velocity of an identically initialized impermeable vesicle
(green). The bottom right plot is the maximum norm of the different
velocity contributions as a function of time. The black vertical lines
correspond to the times reported in other three subplots. At $t=0.5$, semi-permeability velocity is small relative to the
force balance velocity, and the semi-permeable velocity matches with the
impermeable vesicle velocity. At time $t=5$, the velocity of the
impermeable vesicle, and also the force balance velocity, are comparable
in size to the semi-permeability velocity, and the semi-permeable
vesicle begins to inflate. Then, at time $t=50$, the impermeable vesicle
is at its steady state, while the semi-permeable vesicle still has
contributions due to semi-permeability and force balance. We see that
the velocity of a semi-permeable vesicle initially agrees with the
velocity of an impermeable vesicle. Then, as the force balance velocity
approaches zero, the semi-permeable velocity becomes significant,
causing the vesicle to inflate until it reaches a steady state circular
shape.

\begin{figure}[htp]
  \centering
  \includegraphics[width=0.48\linewidth]{figures/ellipseVelocity1.pdf}
  \includegraphics[width=0.48\linewidth]{figures/ellipseVelocity2.pdf}
  \includegraphics[width=0.48\linewidth]{figures/ellipseVelocity3.pdf}
  \includegraphics[width=0.48\linewidth]{figures/ellipseVelocityNorms.pdf}
  \caption{\label{fig:vesVelocity} The velocity contributions due to
  force balance (blue) and semi-pemeability (red). The green dashed line
  is the velocity of an impermeable vesicle with the same initial shape
  as the semi-permeable vesicle. The bottom right plot is the max norm
  of each of term that contributes to the vesicle velocity.  The three
  black lines correspond to the times reported in the other three
  plots.}
\end{figure}

Finally, we consider a semi-permeable vesicle, with $\beta=1$, that
initially has regions with positive curvature and others with negative
curvature. The vesicle tension, $\Lambda$, and the flux $\ff \cdot \nn$,
are plotted in Figure~\ref{fig:starTensionFlux}. The vesicle undergoes
three general behaviors: (i) For $t \in (0,1.6 \times 10^{-2})$, the
vesicle has a net negative flux resulting in a loss of area; (ii) For $t
\in (1.6 \times 10^{-2},1.0)$, the vesicle has a net positive flux and
the tension most of the membrane is undergoing negative tension; (iii)
For $t \in (1.0,2.0)$, the vesicle is near its steady state and
transitions to a positive constant tension. In the first time interval,
the vesicle deflates. Near the points of largest negative curvature,
there is a large negative tension and large inflow (positive flux).
Correspondingly, near the points with largest positive curvature, there
is a small tension and outflow (negative flux). In the second time
interval, the tension is negative and the flux is positive nearly
everywhere. The regions with largest negative curvature correspond to
regions with largest negative tension and the most inflow. Finally, in
the last time interval, the tension transitions to a positive constant
function, the membrane force and flux transition to zero, resulting in a
steady state circular vesicle with positive constant tension.

\begin{figure}[htp]
  \centering
  \includegraphics[width=0.19\linewidth,trim =2cm 5cm 0cm 5cm, clip=true]{figures/StarTensionTime1.pdf}
  \includegraphics[width=0.19\linewidth,trim =2cm 5cm 0cm 5cm, clip=true]{figures/StarTensionTime2.pdf}
  \includegraphics[width=0.19\linewidth,trim =2cm 5cm 0cm 5cm, clip=true]{figures/StarTensionTime3.pdf}
  \includegraphics[width=0.19\linewidth,trim =2cm 5cm 0cm 5cm, clip=true]{figures/StarTensionTime4.pdf}
  \includegraphics[width=0.19\linewidth,trim =2cm 5cm 0cm 5cm,clip=true]{figures/StarTensionTime5.pdf}

  \includegraphics[width=0.19\linewidth,trim =2cm 5cm 0cm 5cm, clip=true]{figures/StarFluxTime1.pdf}
  \includegraphics[width=0.19\linewidth,trim =2cm 5cm 0cm 5cm, clip=true]{figures/StarFluxTime2.pdf}
  \includegraphics[width=0.19\linewidth,trim =2cm 5cm 0cm 5cm, clip=true]{figures/StarFluxTime3.pdf}
  \includegraphics[width=0.19\linewidth,trim =2cm 5cm 0cm 5cm, clip=true]{figures/StarFluxTime4.pdf}
  \includegraphics[width=0.19\linewidth,trim =2cm 5cm 0cm 5cm, clip=true]{figures/StarFluxTime5.pdf}
  \caption{\label{fig:starTensionFlux} The vesicle tension (top) and
  flux (bottom) of the semi-permeable vesicle in a quiescent
  flow with $\beta=1$.}
\end{figure}

@@

>>>>

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section*{Examples}
We consider a single semi-permeable vesicle in four different flow
regimes. All vesicles have a length of $L=4\pi$, but the area is not
conserved because of semi-permeability. However, the area is bounded by
$4\pi$ which is the area of a circle with length $L=4\pi$. For each
simulation, we apply an SDC second-order adaptive time stepping
method~\cite{qua-bir2016}.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection*{Quiescent Flow} 
We consider a semi-permeable elliptical vesicle with initial reduced
area $\nu = 0.65$ and several different permeability rates in a
quiescent flow. The steady-state of an impermeable vesicle has constant
negative tension~\cite{kra-win-sei-lip1996}, but based
on~\eqref{eqn:SSshape}, a semi-permeable vesicle relaxes to a circle
with positive constant tension $\Lambda = 2\pi^2/L^2$. Recalling the
governing equation for the area in~\eqref{eqn:areaROC}, we expect both
the time for the vesicle to deviate from its initial area, and the time
to reach its steady state will scale with $\beta^{-1}$. In
Table~\ref{tbl:ellipseRelaxTimes}, we report the time,
$t_\mathrm{steady}$, required for the vesicle's area to be within 1.3\%
of its maximum area, and the time, $t_\mathrm{inflate}$, required for
the vesicle's area to exceed 1.3\% of its initial area. We see that both
times scale with $\beta^{-1}$. We also plot the area of a vesicle with
six different permeability rates in Figure~\ref{fig:ellipseArea}. The
area is computed directly from the vesicle shape (solid lines) and by
integrating equation~\eqref{eqn:areaROC} (marks) with a first-order
method. We also include black dashed lines indicating the times,
$t_{\mathrm{steady}}$, that the vesicle is near its steady state
circular shape.
\begin{table}[htp]
  \centering
  \caption{\label{tbl:ellipseRelaxTimes} The time for a semi-permeable
  vesicle to its steady state ($t_\mathrm{steady}$) and to begin
  inflating ($t_\mathrm{inflate}$).}
  \begin{tabular}{C{0.8cm}C{2cm}C{0.8cm}C{2cm}C{0.8cm}}
    $\beta$ & $t_\mathrm{steady}$ & ratio & $t_\mathrm{inflate}$ & ratio \\
    \midrule
    $10^{0}$  & $6.45 \times 10^{0}$ & ---  
              & $2.17 \times 10^{-2}$ & --- \\ 
    $10^{-1}$ & $2.00 \times 10^{1}$ & 3.10 & 
                $1.25 \times 10^{-1}$ & 5.76 \\
    $10^{-2}$ & $9.61 \times 10^{1}$ & 4.81 & 
                $1.59 \times 10^{0}$ & 12.72 \\
    $10^{-3}$ & $8.36 \times 10^{2}$ & 8.71 & 
                $2.19 \times 10^{1}$ & 13.77 \\
    $10^{-4}$ & $8.15 \times 10^{3}$ & 9.74 & 
                $2.14 \times 10^{2}$ & 9.77 \\
    $10^{-5}$ & $8.03 \times 10^{4}$ & 9.85 & 
                $2.20 \times 10^{3}$ & 10.14 \\
    \bottomrule
  \end{tabular}
\end{table}

@@

We further analyze the transient dynamics for small $\beta$ by
considering the vesicle velocity~\eqref{eqn:vesVelocity} due to
semi-permeability, $\beta(\ff \cdot \nn)\nn$, and force balance,
$\SS[\ff]$. Initially, since $\beta$ is small, the semi-permeability
velocity is much smaller than the force balance velocity. Therefore, the
velocity resembles that of an impermeable vesicle. As the
vesicle approaches the steady state shape of an impermeable vesicle, the
force balance velocity decreases below the semi-permeability velocity,
and then the semi-permeable vesicle begins to dominate, causing the
vesicle to inflate.  As the vesicle becomes more circular, the
membrane force and the corresponding velocity both converge to
zero. In Figure~\ref{fig:vesVelocity} we plot the force balance velocity
(blue) and the semi-permeability velocity (red) at three times with
$\beta=10^{-3}$. We also plot the velocity of an identically initialized
impermeable vesicle (green).  At $t=0.5$, semi-permeability velocity is
small relative to the force balance velocity, and the semi-permeable
velocity matches with the impermeable vesicle velocity. At time $t=5$,
the magnitude of the velocity of the impermeable vesicle and the force
balance velocity are both comparable to the magnitude of the
semi-permeable velocity, and the semi-permeable vesicle begins to
inflate. Then, at time $t=50$, the impermeable vesicle is at its steady
state, while the semi-permeable vesicle still has contributions due to
semi-permeability and force balance. The bottom right plot is the
maximum norm of the different velocity contributions as a function of
time, and the black vertical lines correspond to the times reported in
other three subplots.

%  \input{figures/ellipseVelocity1.tikz}
%  \input{figures/ellipseVelocity2.tikz}
%  \input{figures/ellipseVelocity3.tikz}
%  \input{figures/ellipseVelocityNorms.tikz}
\begin{figure}[htp]
  \centering
%  \input{figures/ellipseVelocity1.tikz}
%  \input{figures/ellipseVelocity2.tikz}
%  \input{figures/ellipseVelocity3.tikz}
%  \input{figures/ellipseVelocityNorms.tikz}
  \includegraphics[width=0.48\linewidth]{figures/ellipseVelocity1.pdf}
  \includegraphics[width=0.48\linewidth]{figures/ellipseVelocity2.pdf}
  \includegraphics[width=0.48\linewidth]{figures/ellipseVelocity3.pdf}
  \includegraphics[width=0.48\linewidth]{figures/ellipseVelocityNorms.pdf}
  \caption{\label{fig:vesVelocity} The velocity contributions due to
  force balance (blue) and semi-pemeability (red). The green dashed line
  is the velocity of an impermeable vesicle with the same initial shape
  as the semi-permeable vesicle. The bottom right plot is the max norm
  of each of term that contributes to the vesicle velocity.  The three
  black lines correspond to the times reported in the other three
  plots.}
\end{figure}

\todo[inline]{redo this paragraph}
Finally, letting $\beta=1$, we consider a semi-permeable vesicle that
initially has regions of positive and negative curvature. The vesicle
tension, $\Lambda$, and the flux, $\ff \cdot \nn$,
are plotted in Figure~\ref{fig:starTensionFlux}. The vesicle undergoes
three general behaviors: (i) For $t \in (0,1.6 \times 10^{-2})$, the
vesicle has a net negative flux resulting in a loss of area; (ii) For $t
\in (1.6 \times 10^{-2},1.0)$, the vesicle has a net positive flux and
the tension most of the membrane is undergoing negative tension; (iii)
For $t \in (1.0,2.0)$, the vesicle is near its steady state and
transitions to a positive constant tension. In the first time interval,
the vesicle deflates. Near the points of largest negative curvature,
there is a large negative tension and large inflow (positive flux).
Correspondingly, near the points with largest positive curvature, there
is a small tension and outflow (negative flux). In the second time
interval, the tension is negative and the flux is positive nearly
everywhere. The regions with largest negative curvature correspond to
regions with largest negative tension and the most inflow. Finally, in
the last time interval, the tension transitions to a positive constant
function, the membrane force and flux transition to zero, resulting in a
steady state circular vesicle with positive constant tension.

\begin{figure}[htp]
  \centering
  \includegraphics[width=0.18\linewidth,trim =2cm 5cm 0cm 5cm, clip=true]{figures/StarTensionTime1.pdf}
  \includegraphics[width=0.18\linewidth,trim =2cm 5cm 0cm 5cm, clip=true]{figures/StarTensionTime2.pdf}
  \includegraphics[width=0.18\linewidth,trim =2cm 5cm 0cm 5cm, clip=true]{figures/StarTensionTime3.pdf}
  \includegraphics[width=0.18\linewidth,trim =2cm 5cm 0cm 5cm, clip=true]{figures/StarTensionTime4.pdf}
  \includegraphics[width=0.18\linewidth,trim =2cm 5cm 0cm 5cm,clip=true]{figures/StarTensionTime5.pdf}

  \includegraphics[width=0.18\linewidth,trim =2cm 5cm 0cm 5cm, clip=true]{figures/StarFluxTime1.pdf}
  \includegraphics[width=0.18\linewidth,trim =2cm 5cm 0cm 5cm, clip=true]{figures/StarFluxTime2.pdf}
  \includegraphics[width=0.18\linewidth,trim =2cm 5cm 0cm 5cm, clip=true]{figures/StarFluxTime3.pdf}
  \includegraphics[width=0.18\linewidth,trim =2cm 5cm 0cm 5cm, clip=true]{figures/StarFluxTime4.pdf}
  \includegraphics[width=0.18\linewidth,trim =2cm 5cm 0cm 5cm, clip=true]{figures/StarFluxTime5.pdf}
  \caption{\label{fig:starTensionFlux} The vesicle tension (top) and
  flux (bottom) of the semi-permeable vesicle in a quiescent
  flow with $\beta=1$.}
\end{figure}

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection*{Stenosis}
We consider a semi-permeable vesicle in a constricted geometry. A
double-layer potential is used to impose a boundary condition on the
outer wall, and near-singular integration~\cite{qua-bir2014} quadrature
resolves the vesicle-wall interaction.
